\SW: II. New lens candidates from the CFHTLS discovered through citizen science

### 2015-03-02 doc/sw-cfhtls.tex
rearranged discussion section, added trajectory fig

%magdiff=5*0.1;
%zdiffq=0.1;

@@

\subsection{False negatives: known lenses missed by \sw}
\label{sec:fn}
Like any lens finding method, the \sw system can fail to detect
certain kinds of lenses. We explore and discuss why some of the known
confirmed lenses may have been missed. About 40\% of the known confirmed sample of
lenses are missed (see \Tref{tab:stats}). And, XXXX

Among the confirmed lenses from the \rf, about 50\% are missed. Out of
the missed sample of 18 lenses, about half of them are visually
difficult to detect and the other half seem to show faint blue smudges
around galaxies which should have been easier to identify.  {\bf Let us
compare the two samples to see if there are any obvious differences.
only 2 out of this sample of 18 were detected at stage I (reached high P
even with fewer classifications) and missed at stage II (P value was low
even after recieving very many classifications). The rest of the 16
lenses were rejected right at stage I XXX}

If we consider the \af lenses, only 6 out of the 26 ($\sim$23\%) were
missed by \sw. One of those 6 lenses was detected at stage I but failed
to get high enough P value at stage II even after receiving a total of
50 classifications (set as the upper limit). The rest of the 5 lenses
are already rejected at stage I. Upon visual inspection of these 5
systems, they appear to be difficult to identify as lenses because
either the lensed images are faint, have low curvature, are close to the
edge and/or are not visible due to proximity to an extremely bright
galaxy ({\bf show some examples XXX}).

@@

Efficiency of a visual search can vary in different sections of an
image. Our eyes tend to focus usually at the center of an image and lens
candidates close to the borders could go undetected. Therefore, it is
essential to test and understand if \sw tends to miss certain lenses
because they happen to be close to the borders of the image cutouts.

From the SWAP, the image cutouts inspected by the \sw volunteers receive
a status of detected (if P$>$P$_{\rm acc_thresh}$), rejected (if
P$<$P$_{\rm rej_thresh}$) and undecided (if P$_{\rm rej_thresh}$ $<$ P
$<$ P$_{\rm acc_thresh}$). In \Fref{fig:comppos}, we compare the
positions of simulated lenses which are detected (red), undecided
(green) and rejected (blue). We also make the same comparison for the
known lens candidates in the right panel.  We note that for some cases
randomly selected subsamples are shown for the ease of visual
comparison. For example, the absolute number of simulated lenses
detected is too high compared to the rejected sample. We do not find any
strong visual correlation in the rate of detections as a function of the
position for both the simulated and the known lens sample. Thus, the
completeness of the lens sample does not seem to be significantly
affected by whether a lens is located close to the border or well within
the center.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\subsection{Why \sw candidates were missed by lens finding robots?}

The new lens candidates discovered by \sw are run through the \rf to
trace and understand at what stage the algorithm failed to detect them.

At the first step of the \rf algorithm, a galaxy catalog is generated
based on magnitude, redshift and SED type \citep[see]{Gavazzi2014} to
select galaxies which are most likely to act as lenses. We find that
about 40\% of the new \sw candidates fail to meet this initial selection
criteria e.g. SW1, SW13, SW19, SW22, SW27 and SW30.  All of the lensing
galaxies are bright enough to satisfy the $i<22$ criterion. However,
some of them have a bright companion galaxy, some of them do not look
like E/S0 type galaxies and some are edge on galaxies. {\bf there are a
few which look regular galaxies so which criteria are they failing??}

Next, we rerun the \af on the same \sw sample of new candidates. The \af
is directly run an image to look for elongated arc like objects and does
not require a list of targets to begin with. Objects are identified by
placing thresholds on the noise level in the images. Thus, \af
detections are sensitive to changes in the noise levels. Also, the size
of the PSF is important to know what dimensions should be expected for
the arc candidates.

Originally, the \af was run on a large image with an area of $\sim 19400
\times 19400$ pixels$^2$. For the rerun, we work with much smaller
images because this is faster but this affects the number and type of
arc detections. We find that about 30\% of the new candidates are detected
without changing any of the thresholds in the code because of the change
in the noise level.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\subsection{Limitations and Caveats of the training sample}

The simulated sample is not perfect because our understanding of various
phenomena in the Universe is not complete and because of the uncertainties in
the parameters of our model. Here, we describe some of the cases or
aspects in which the simulations are known to have failed or seem
unrealistic.

We use single Sersic component to describe the light profiles of background
galaxies. This is clearly not the most accurate description for galaxies,
especially, star-forming galaxies which form a significant fraction of the
lensed galaxy population. Star-forming galaxies have complex structures such as
star forming knots, spiral arms, bars and disks.

@@

>>>>

\subsection{Limitations of the simulated sample}
The simulated sample is not perfect because our understanding of various
phenomena in the Universe is not complete and because of the uncertainties in
the parameters of our model. Here, we describe some of the cases or
aspects in which the simulations are known to have failed or seem
unrealistic.

We use single Sersic component to describe the light profiles of background
galaxies. This is clearly not the most accurate description for galaxies,
especially, star-forming galaxies which form a significant fraction of the
lensed galaxy population. Star-forming galaxies have complex structures such as
star forming knots, spiral arms, bars and disks. 
%magdiff=5*0.1;
%zdiffq=0.1;

@@

\subsection{Why \sw candidates were missed by lens finding robots?}
We rerun the \rf and \af on images centered on the new \sw candidates to
trace and understand at what stage the algorithm failed to detect them.

First, we rerun \rf on the new \sw sample. At the beginning, a galaxy
catalog is generated based on magnitude, redshift and SED type
\citep[see]{Gavazzi2014} to select galaxies which are most likely to act
as lenses. We find that about 40\% of the new \sw candidates fail to
meet this initial selection criteria e.g. SW1, SW13, SW19, SW22, SW27
and SW30.  All of the lensing galaxies are bright enough to satisfy the
$i<22$ criterion. However, some of them have a bright companion galaxy,
some of them do not look like E/S0 type galaxies and some are edge on
galaxies. {\bf there are a few which look regular galaxies so which
criteria are they failing??}

Next, we rerun the \af on the same \sw sample of new candidates. The \af
is directly run on images to look for elongated arc like objects and does
not require a list of targets to begin with. Objects are identified by
placing thresholds on the noise level in the images. Thus, \af
detections are sensitive to changes in the noise levels.

Originally, the \af was run on a large image with an area of $\sim 19350
\times 19350$ pixels$^2$. For the rerun, we work with much smaller
images because this is faster but this alters the measured
noise and hence, affects the number and type of arc detections. We find
that about 30\% of the new candidates are detected without changing any
of the thresholds in the code because of the change in the noise level.
{\bf does this make sense?}

The \af code calculates second order brightness moments around every
pixel to decide if the distribution of flux is elongated in some
direction in order to detect elongated arc-like objects. An elongation
estimator is assigned to every pixel. All pixels with a value of the
estimator above a certain threshold are connected to form the arc
feature. This is building of the arc candidate. Subsequently, arc
properties such as the area, mean flux, length and curvature are
determined. We relax the threshold at the building stage and also relax
thresholds mainly on the area of the arc which leads to detection of
about 75\% of the new \sw candidates. We find that relaxing thresholds
on other arc properties do not improve the detection rate significantly
and thus are left unchanged.

Typically, the candidates are missed from the \af sample if a) they are
brighter in the redder band rather than the bluer $g$ band used for arc
detection because then the arc feature will be at the noise level and
will fail detection b) the flux of the arc and the galaxy are
blended in the $g$ band such that the \af mistakenly connects part of the
galaxy to the pixels belonging to the arc candidate which then results in a
candidate with no morphological characteristics of an arc or c) if the
candidate is not elongated which means most of the lensed quasars with
circular looking lensed images will be missed.

Relaxing the thresholds obviously increases the total number of
candidate detections which includes a large sample of false positives.
For example, the number of arc candidate detections increase by a factor
of $\sim$2 when we relaxed the thresholds in the reruns described above
whereas the number of false positives increase by a factor of $\sim$5.
Thus, it is not recommended to relax these thresholds alone. A better
alternative is to relax the thresholds to increase the completeness and
cross-correlate the arc candidate positions with a galaxy catalog to
discard those candidates which are not close to a putative lensing
galaxy within typical radius in order to reduce the false positives.

%% 19,46 (original and relaxed criteria); 72,338 (smaller image); 1548,7845 (larger image)

\subsection{False negatives: known lenses missed by \sw}
\label{sec:fn}
Like any lens finding method, the \sw system can fail to detect certain
kinds of lenses. Here, we explore and discuss why some of the known
lenses may have been missed.

We find that about 40\% of the known confirmed sample of lenses are
missed (see \Tref{tab:stats}). While this seems like a fraction, we note
that many of the missed lenses are from the \rf sample with small
Einstein radii and faint lensed images (see \Fref{fig:compre}). Among
the confirmed lenses from the \rf, about 50\% are missed. Out of the
missed sample of 18 lenses, about half of them are visually difficult to
detect and the other half appear to have faint blue smudges around
galaxies which should have been easier to identify. Similarly,
if we consider the \af lens sample, $\sim$23\% are missed by \sw. This
is a relatively small sample of $\sim 5$ systems and visual inspection
suggests that, by and large, either the lensed features are faint
or they have odd properties which makes them difficult to identify
correctly.  ({\bf show some
example images XXX}). For further tests, we combine the \rf and \af sample.

For a lens finding method which uses collective skill, experience and
knowledge of a group of citizens, it may be difficult to find a single
factor with certainty which causes a lens candidate to be missed. We
attempt to understand whether there is indeed a single dominant factor
that is resulting in the loss of these lenses or the lenses are being
missed due to a combination of multiple reasons. Below, we consider some
of the factors that could affect the efficiency of finding lenses.

First, we check if the Nclass is different for the detected and the
missed sample. Surprisingly, most of the lenses in the known sample have
few classifications (Nclass<10) which includes both the detected and
missed lenses. Many of the remaining have between 10 to 20
classifications. And, only a few subjects have Nclass>20 which continue
to remain for long in the database because these candidates may have
been difficult to identify.

%{\bf Let us compare the two samples to see if there are any obvious differences. only 2 out of this sample of 18 were detected at stage I (reached high P even with fewer classifications) and missed at stage II (P value was low even after recieving very many classifications). The rest of the 16 lenses were rejected right at stage I XXX}
%One of those 6 lenses was detected at stage I but failed to get high enough P value at stage II even after receiving a total of 50 classifications (set as the upper limit). The rest of the 5 lenses are already rejected at stage I. Upon visual inspection of these 5 systems, they appear to be difficult to identify as lenses because either the lensed images are faint, have low curvature, are close to the edge and/or are not visible due to proximity to an extremely bright galaxy

@@

\subsubsection{Blind lens search}
Efficiency of a visual search can vary in different sections of an
image. Our eyes tend to focus usually at the center of an image and lens
candidates close to the borders could go undetected. Therefore, it is
essential to test and understand if \sw tends to miss certain lenses
because they happen to be close to the borders of the image cutouts.

From the SWAP, the image cutouts inspected by the \sw volunteers receive
a status of detected (if P$>$P$_{\rm acc thresh}$), rejected (if
P$<$P$_{\rm rej thresh}$) and undecided (if P$_{\rm rej thresh}$ $<$ P
$<$ P$_{\rm acc thresh}$). In \Fref{fig:comppos}, we compare the
positions of simulated lenses which are detected (red), undecided
(green) and rejected (blue). We also make the same comparison for the
known lens candidates in the right panel.  We note that for some cases
randomly selected subsamples are shown for the ease of visual
comparison. For example, the absolute number of simulated lenses
detected is too high compared to the rejected sample. We do not find any
strong visual correlation in the rate of detections as a function of the
position for both the simulated and the known lens sample. Thus, the
completeness of the lens sample does not seem to be significantly
affected by whether a lens is located close to the border or well within
the center.

\subsubsection{Volunteer profile}

Here, we investigate if the skill or experience of volunteers is
systematically different between the sample of detected and missed
lenses.

We check how the posterior probability P (see XXXX Paper I) of an image
or a subject to contain a lens changes as the image receives more
classifications from multiple users.  In \Fref{fig:detmis}, we show the
trajectory plots of a few examples of detected lenses (top left panels)
and missed lenses (bottom left panels) by \sw at stage I. The number of
classifications (Nclass) for a subject increase from top to bottom.
Also, every subject is assigned a prior probability P0=2.e-4 (grey
dashed line) and starts at the middle of the trajectory plot. The P
value of a subject is updated with every classification from the
volunteer.  If a volunteer identifies a lens candidate, the trajectory
moves to the right otherwise moves to the left. A subject is accepted if it
crosses the blue-dashed line marking the (P$_{\rm acc thresh}=0.95$) on
the right and is rejected, if it crosses the red-dashed line marking the
P$_{\rm rej thresh}=1.e-7$ on the left.

By how much the P value will change depends on how well the
volunteers are performing on the training sample. Thus, more skilled
volunteers will change the P by a large factor compared to the less
skilled volunteers. This is evident in the trajectory plots as large and
small distances in between consequent points.

@@

\begin{figure*}
\begin{center}
\includegraphics[scale=0.6]{sw-cfhtls-figs/det_mis_traj.pdf}
\includegraphics[scale=0.6]{sw-cfhtls-figs/det_mis_hist.pdf}
\caption{ \label{fig:detmis}
Example cases of trajectories of known lenses on the left and histogram
of the known lenses. The trajectories on the left are lenses that are
detected by \sw are at the top and lenses that are missed are at the bottom. }
\end{center}
\end{figure*}
